Collaboratively Testing the Validity of Neuroscientific Models

### 2014-03-21 neurounit_frontiers/Paper.tex
Changed NeurounitFrontiers to neurounit_frontiers

  def _judge(self, model):
    inputs, means, stds = self.inputs, self.means, self.stds
    n = len(inputs)
    counts = numpy.empty((n,))
    for i in xrange(n):
      counts[i] = model.spike_count_from_current(inputs[i])
    chisquared = sum((counts-means)**2 / means) # An array of chi-squared values.  
    p = scipy.stats.chi2.cdf(chisquared,n-1) # An array of p-values.  
    pooled_p = sciunit.utils.fisherp(p_array) # A pooled p-value.  
    return sciunit.PValue(pooled_p, related_data={
      "inputs": inputs, "counts": counts, "empirical_means": means, "empirical_stds": stds
      })
\end{python}
\vspace{-15px}
\caption{\texttt{[neuronunit.SpikeCountTest]} An example single neuron spike count test class implemented using \textit{SciUnit}. Because this implementation contains logic common to many different systems, \textit{NeuronUnit} was developed to provide a simpler means to deliver it (see Sec. \ref{sec:neuronunit}).}
\label{fig:rate_test}
\vspace{-10px}
\end{figure}

@@

\begin{figure}
\begin{python}
class SpikeCountFromCurrent(sciunit.Capability):
  def spike_count_from_current(self, input): 
    """Takes a numpy array containing current stimulus (in nA) and
    produces an integer spike count. Can be called multiple times."""
    raise NotImplementedError("Model does not implement capability.")
\end{python}
\vspace{-15px}
\caption{\texttt{[neuronunit.SpikeCountFromCurrent]} An example capability specifying a single required method (used by the test in Figure \ref{fig:rate_test}).}
\label{fig:capability}
\vspace{-10px}
\end{figure}
\begin{figure}
\begin{python}
class TrainSpikeCountFromCurrent(sciunit.Capability):
  def train_with_currents(self, currents, counts):
    """Takes a list of numpy arrays containing current stimulus (in nA) and
    observed spike counts. Model parameters should be adjusted based on this
    training data."""
    raise NotImplementedError("Model does not implement capability.")
\end{python}
\vspace{-15px}
\caption{\texttt{[neuronunit.TrainSpikeCountFromCurrent]} Another capability specifying a training protocol (not used by the test in Figure \ref{fig:rate_test}).}
\label{fig:training}
\vspace{-10px}
\end{figure}

@@

  def train_with_currents(self, currents, counts):
    means = [numpy.mean(c) for c in currents]
    [self.offset, self.scale] = numpy.polyfit(means, counts, deg=1)    
\end{python}
\vspace{-15px}
\caption{\texttt{[neuronunit.LinearModel]} A model that returns a spike count by applying a linear transformation to the mean input current. The parameters can be provided manually or learned from data provided by a test or user (see text).}
\label{fig:simple_model}
\vspace{-10px}
\end{figure}

@@

\end{figure}
The NeuroElectro project (\url{http://neuroelectro.org}) is an effort to make all published data on single cell neurophysiology available in a machine-readable format \citep{tripathy_neuroelectro:_2012}.  
Currently, up to 27 electrophysiological properties are available for 93 cell types, gathered from over 2000 single pieces of published data extracted from tables in published papers. 
We have made it easy to parameterize the test families in \textit{Neuron\-Unit}  using data extracted from the NeuroElectro API. 
Tests can be based upon data from single journal articles, or from ensembles of articles with a common theme (e.g. about a particular neuron type). 
The latter is illustrated in Figure \ref{fig:neuronunit_example}, on lines 1-4. 
Once data has been retrieved from NeuroElectro, associated statistics (e.g. mean, standard error and sample size) can be extracted to parameterize tests (e.g. a test of spike widths on line 10). 
Data from NeuroElectro can be used to validate a variety of basic electrophysiological features of neuron models, including spike width, resting membrane potential, after-hyperpolarization amplitude, etc. \todo{write a sentence or two about its use of the NeuroLex repository}
As NeuroElectro is the only public, curated source of such data, it represents a key resource facilitating test construction.

@@

NeuroConstruct (\cite{gleeson_neuroconstruct:_2007}, \url{http://www.neuroconstruct.org}) is a simulation manager that takes NeuroML models and hands them off to a supported simulator for execution. A number of popular simulators support execution of NeuroML models, including NEURON, GENESIS (\cite{bower_genesis_2007}, \url{http://genesis-sim.org}), NEST (\cite{gewaltig_nest_2007}, \url{http://www.nest-initiative.org}), and MOOSE (\cite{ray_moose_2008}, \url{http://moose.ncbs.res.in}). 
\textit{NeuronUnit} provides a \verbx{sciunit.Model} subclass called \verbx{NeuroConstructModel} that takes a path to a NeuroML model as a parameter.  
NeuroML can describe a wide range of models, but our \verbx{NeuroConstructModel} class currently supports only time-integrable neuron models with membrane potentials (exposing capabilities \verbx{TimeIntegrable} and \verbx{HasMembranePotential} at minimum; see Sec. \ref{sec:neurotools}). 
It can be instantiated directly or subclassed (useful when additional capabilities need to be manually exposed, for example) to generate a model suitable for testing (Fig. \ref{fig:ca1_model}). The underlying simulator can be configured (defaulting to ??\todo{What does it default to in our code??}; not shown). In this example, we simply create a model based on a NeuroML model of CA1 cells downloaded from OpenSourceBrain to the local file system (\url{http://opensourcebrain.org/projects/ca1pyramidalcell}).\todo{Migliore ref.}\todo{Should OSB models be transiently downloaded as part of the instantiation of OSBModel?}

@@

To directly test a model in OSB, we can skip the step of downloading it the local file system or adding it to the test repository and simply download it from OSB directly by using the \verbx{OSBModel} class, shown on lines 5-8 of Figure \ref{fig:neuronunit_example}.

@@

\begin{figure}
\begin{python}
class CA1PyramidalCellModel(NeuroConstructModel):
	"""CA1 Pyramidal Cell model in the local file system."""
	def __init__(self):
		project_path = neuroconstruct.get_path("hippocampus","CA1_pyramidal_neuron","CA1PyramidalCell","neuroConstruct")
		NeuroConstructModel.__init__(self,project_path)
\end{python}
\vspace{-15px}
\caption{A model class corresponding to a CA1 Pyramidal Cell model downloaded locally from Open Source Brain}
\label{fig:ca1_model}
\vspace{-10px}
\end{figure}

@@

We use NeuroTools to implement a variety of \textit{SciUnit} capabilities in \textit{NeuronUnit}. 
For example, an \verbx{AnalogSignal} object (e.g. a membrane potential time series) from NeuroTools has a threshold detection method that returns a NeuroTools \verbx{SpikeTrain} object. 
The \textit{NeuronUnit} capability \verbx{HasSpikeTrain}  requires that a method named \verbx{get_spike_train} be implemented, returning a \verbx{SpikeTrain} object. 
\verbx{NeuroConstructModel} implements this by default by calling the NeuroTools method \verbx{AnalogSignal.threshold_detection}, retrieving the membrane potential by calling the \verbx{get_membrane_potential} method implemented by the \verbx{HasMembranePotential} capability. 
This capability is implemented by the OSBModel class that supports all current OSB models.
NeuroTools is used in this manner to implement a wide variety of capabilities for \verbx{NeuroConstructModel}s, without requiring that the modeler do so explicitly (though they can override these if they wish). 
This also helps scientists by directing them to a set of common utility functions and data object types that they can use, knowing that they are accepted by the community around NeuronUnit.\todo{do you have a list of capabilities that neuroconstructmodels can automatically implement?}

@@

\subsection{A Complete Pipeline}
Although the tools described in Sec. \ref{sec:neuronunit} do not exhaust the possible sources of models, capabilities, and test data, they provide an immediate point of entry into the neurophysiology community and a powerful demonstration of our proposal. 
In the \textit{hippocampus\_tests} repository (\url{http://github.com/scidash/hippocampus\_tests}) is an IPython notebook (\textit{ca1\_suite.ipynb}) demonstrating a complete test suite similar to the \emph{ad hoc} suite used by the QSNMC. 
Our tests are parameterized by data drawn both from NeuroElectro and entered manually, as well as models drawn both from OSB and implemented manually. 
New tests, data and models can be added by contributing to these community resources and/or submitting pull requests; 
we anticipate handing control over merging these requests to members of the hippocampal modeling community once our interface stabilizes and this paper is published. 
This repository represents a template that other groups of systems neuroscientists can use for their system of interest, and a starting point for other research areas in neuroscience, and science more broadly.

@@

\subsection{Diversity of Levels and Kinds of Models and Data}
The diversity of topics in biology is vast. 
\textbf{First}, we address this by providing an interface allowing modelers to flexibly express precisely how a model should be interfaced using capabilities. 
Capabilities can inherit from and call into each other. 
This is more flexible than the fixed interfaces used in existing modeling competitions or domains like machine learning where only a few interfaces are of interest (e.g. classifiers). 
\textbf{Second}, NeuroML naturally addresses diversity of scales because it is organized hierarchically, in ``levels.''  
Models can be sub- or supersets of other models; similarly for SBML (\cite{hucka_systems_2003}, \url{http://sbml.org}) a general systems biology markup language. 
\textbf{Third}, cross-level testing can use ``Representional Similarity Analysis'' (RSA) \citep{kriegeskorte_representational_2008}, requiring only that a model respond to defined inputs (e.g. stimuli)\todo{I have no idea what this means}. 
A ``similarity matrix'' for input responses defines a unique model signature, and can serve as intermediate test output. 
Goodness-of-fit between similarity matrices for model and experiment determines test scores; 
these matrices are independent of model scale because their size depends only on test inputs, not system detail.

>>>>

  def _judge(self, model):
    inputs, means, stds = self.inputs, self.means, self.stds
    n = len(inputs)
    counts = numpy.empty((n,))
    for i in xrange(n):
      counts[i] = model.spike_count_from_current(inputs[i])
    chisquared = sum((counts-means)**2 / means) # An array of chi-squared values.  
    p = scipy.stats.chi2.cdf(chisquared,n-1) # An array of p-values.  
    pooled_p = sciunit.utils.fisherp(p_array) # A pooled p-value.  
    return sciunit.PValue(pooled_p, related_data={
      "inputs": inputs, "counts": counts, "empirical_means": means, "empirical_stds": stds
      })
\end{python}
\vspace{-15px}
\caption{\texttt{[SpikeCountTest]} An example single neuron spike count test class implemented using \textit{SciUnit}. Because this implementation contains logic common to many different systems, \textit{NeuronUnit} was developed to provide a simpler means to deliver it (see Sec. \ref{sec:neuronunit}).}
\label{fig:rate_test}
\vspace{-10px}
\end{figure}

@@

\begin{figure}
\begin{python}
class SpikeCountFromCurrent(sciunit.Capability):
  def spike_count_from_current(self, input): 
    """Takes a numpy array containing current stimulus (in nA) and
    produces an integer spike count. Can be called multiple times."""
    raise NotImplementedError("Model does not implement capability.")
\end{python}
\vspace{-15px}
\caption{\texttt{[SpikeCountFromCurrent]} An example capability specifying a single required method (used by the test in Figure \ref{fig:rate_test}).}
\label{fig:capability}
\vspace{-10px}
\end{figure}
\begin{figure}
\begin{python}
class TrainSpikeCountFromCurrent(sciunit.Capability):
  def train_with_currents(self, currents, counts):
    """Takes a list of numpy arrays containing current stimulus (in nA) and
    observed spike counts. Model parameters should be adjusted based on this
    training data."""
    raise NotImplementedError("Model does not implement capability.")
\end{python}
\vspace{-15px}
\caption{\texttt{[TrainSpikeCountFromCurrent]} Another capability specifying a training protocol (not used by the test in Figure \ref{fig:rate_test}).}
\label{fig:training}
\vspace{-10px}
\end{figure}

@@

  def train_with_currents(self, currents, counts):
    means = [numpy.mean(c) for c in currents]
    [self.offset, self.scale] = numpy.polyfit(means, counts, deg=1)    
\end{python}
\vspace{-15px}
\caption{\texttt{[LinearModel]} A model that returns a spike count by applying a linear transformation to the mean input current. The parameters can be provided manually or learned from data provided by a test or user (see text).}
\label{fig:simple_model}
\vspace{-10px}
\end{figure}

@@

\end{figure}
The NeuroElectro project (\url{http://neuroelectro.org}) is an effort to make all published data on single cell neurophysiology available in a machine-readable format \citep{tripathy_neuroelectro:_2012}.  
Currently, up to 27 electrophysiological properties are available for 93 cell types, gathered from over 2000 single pieces of published data extracted from tables in published papers. 
NeuroElectro uses the NeuroLex.org ontology \cite{larson_neurolex_2013} to identify specific neuron types and specific electrophysiological properties, and indexes corresponding data values from the literature.  
We have made it easy to parameterize the test families in \textit{Neuron\-Unit}  using data extracted from the NeuroElectro API. 
Tests can be based upon data from single journal articles, or from ensembles of articles with a common theme (e.g. about a particular neuron type). 
The latter is illustrated in Figure \ref{fig:neuronunit_example}, on lines 1-4. 
Once data has been retrieved from NeuroElectro, associated statistics (e.g. mean, standard error and sample size) can be extracted to parameterize tests (e.g. a test of spike widths on line 10). 
Data from NeuroElectro can be used to validate a variety of basic electrophysiological features of neuron models, including spike width, resting membrane potential, after-hyperpolarization amplitude, etc. 
As NeuroElectro is the only public, curated source of such data, it represents a key resource facilitating test construction.

@@

NeuroConstruct (\cite{gleeson_neuroconstruct:_2007}, \url{http://www.neuroconstruct.org}) is a simulation manager that takes NeuroML models and hands them off to a supported simulator for execution. A number of popular simulators support execution of NeuroML models, including NEURON, GENESIS (\cite{bower_genesis_2007}, \url{http://genesis-sim.org}), NEST (\cite{gewaltig_nest_2007}, \url{http://www.nest-initiative.org}), and MOOSE (\cite{ray_moose_2008}, \url{http://moose.ncbs.res.in}). 
\textit{NeuronUnit} provides a \verbx{sciunit.Model} subclass called \verbx{NeuroConstructModel} that takes a path to a NeuroML model as a parameter.  
NeuroML can describe a wide range of models, but our \verbx{NeuroConstructModel} class currently supports only time-integrable neuron models with membrane potentials (exposing capabilities \verbx{TimeIntegrable} and \verbx{HasMembranePotential} at minimum; see Sec. \ref{sec:neurotools}). 
It can be instantiated directly or subclassed (useful when additional capabilities need to be manually exposed, for example) to generate a model suitable for testing (Fig. \ref{fig:ca1_model}). The underlying simulator can be configured (defaulting to NEURON; not shown). In this example, we simply create a model based on a NeuroML model of CA1 cells \cite{migliore_ca1_2005} downloaded from OpenSourceBrain to the local file system upon instantiation (\url{http://opensourcebrain.org/projects/ca1pyramidalcell}).

@@

To directly test a model in OSB, we can skip the step of adding it to the test repository and simply download it from OSB directly by using the \verbx{OSBModel} class, which is a subclass of  \verbx{NeuroConstructModel}, shown on lines 5-8 of Figure \ref{fig:neuronunit_example}.

@@

\begin{figure}
\begin{python}
class CA1PyramidalCellModel(NeuroConstructModel):
	"""CA1 Pyramidal Cell model in the local file system."""
	def __init__(self):
		project_path = neuroconstruct.get_path("hippocampus","CA1_pyramidal_neuron","CA1PyramidalCell","neuroConstruct")
		NeuroConstructModel.__init__(self,project_path)
\end{python}
\vspace{-15px}
\caption{A model class corresponding to a CA1 Pyramidal Cell model (Migliore et al, 2005) downloaded locally from Open Source Brain}
\label{fig:ca1_model}
\vspace{-10px}
\end{figure}

@@

We use NeuroTools to implement a variety of \textit{SciUnit} capabilities in \textit{NeuronUnit}. 
For example, an \verbx{AnalogSignal} object (e.g. a membrane potential time series) from NeuroTools has a threshold detection method that returns a NeuroTools \verbx{SpikeTrain} object. 
The \textit{NeuronUnit} capability \verbx{HasSpikeTrain}  requires that a method named \verbx{get_spike_train} be implemented, returning a \verbx{SpikeTrain} object. 
\verbx{NeuroConstructModel} implements this by default by calling the NeuroTools method \verbx{AnalogSignal.threshold_detection}, retrieving the membrane potential by calling the \verbx{get_membrane_potential} method implemented by the \verbx{HasMembranePotential} capability. 
This capability is implemented by the \verbx{NeuroConstructModel} class which supports all current OSB models and implicitly supports all NeuroML models with somatic membrane potentials.
NeuroTools is used in this manner to implement a wide variety of capabilities for \verbx{NeuroConstructModel}s, without requiring that the modeler do so explicitly (though they can override these if they wish). 
This also helps scientists by directing them to a set of common utility functions and data object types that they can use, knowing that they are accepted by the community around NeuronUnit.

@@

\subsection{A Complete Pipeline}
Although the tools described in Sec. \ref{sec:neuronunit} do not exhaust the possible sources of models, capabilities, and test data, they provide an immediate point of entry into the neurophysiology community and a powerful demonstration of our proposal. 
In the \textit{hippocampus\_tests} repository (\url{http://github.com/scidash/hippocampus\_tests}) we are currently developing a complete test suite similar to the \emph{ad hoc} suite used by the QSNMC, demonstrated by an IPython notebook (\textit{ca1\_suite.ipynb}). 
These tests will be parameterized by data drawn both from NeuroElectro and entered manually, as well as models drawn both from OSB and implemented manually. 
New tests, data and models can be added by contributing to these community resources and/or submitting pull requests; 
we anticipate handing control over merging these requests to members of the hippocampal modeling community once our interface stabilizes and this paper is published. 
This repository represents a template that other groups of systems neuroscientists can use for their system of interest, and a starting point for other research areas in neuroscience, and science more broadly.

@@

\subsection{Diversity of Levels and Kinds of Models and Data}
The diversity of topics in biology is vast. 
\textbf{First}, we address this by providing an interface allowing modelers to flexibly express precisely how a model should be interfaced using capabilities. 
Capabilities can inherit from and call into each other. 
This is more flexible than the fixed interfaces used in existing modeling competitions or domains like machine learning where only a few interfaces are of interest (e.g. classifiers). 
\textbf{Second}, NeuroML naturally addresses diversity of scales because it is organized hierarchically, in ``levels.''  
Models can be sub- or supersets of other models; similarly for SBML (\cite{hucka_systems_2003}, \url{http://sbml.org}) a general systems biology markup language. 
\textbf{Third}, cross-level testing can use ``Representional Similarity Analysis'' (RSA) \citep{kriegeskorte_representational_2008}, requiring only that a model respond to defined inputs (e.g. stimuli). 
In RSA, a ``similarity matrix'' for input responses defines a unique model signature, and can serve as intermediate test output. 
Goodness-of-fit between similarity matrices for model and experiment determines test scores; 
these matrices are independent of model scale because their size depends only on test inputs, not system detail.
